\chapter*{Acknowledgements}
\addcontentsline{toc}{chapter}{Acknowledgements} % Add the preface to the table of contents as a chapter

This thesis recounts part of my journey into the land of formal logic and computation. As 
is customary, the narrative is rather long-winded, and could certainly be 
described as fairly technical at times (it would not look very serious otherwise).
% 
But in the end, the past three years have been much more than a 
bunch of theorems and proofs, and I am grateful to all of you who played 
a part.

Thanks in particular to Nicolas, whom I had the pleasure to have as my 
advisor. For your availability, your insights, your patience, and your 
guidance. I had a lot of fun studying with you during these three years, 
and I learned a lot.

I am also grateful to Ambrus, Conor, Favonia, Jon, Paige, Thierry, and Tom 
for accepting to be part of my jury. It is both very flattering and a bit
intimidating to be heard by so many people I admire.

Honestly, the entire Gallinette team was a blast to interact with during my stay
in Nantes. I had many interesting conversations and pleasant chats with Assia, 
Étienne, Gaëtan, Guihem, Guillaume, Kazuhiko, Kenji, Marie, Matthieu (Piquerez), 
Matthieu (Sozeau), Pierre (Vial), Pierre-Marie and Yannick.
% 
I have particularly fond memories of my interactions with my senior PhD 
students, who taught me the ropes and made me feel at home: 
Ambroise for his humor and cheerfulness,
Simon for making me discover Notre-Dame des Landes and the Jardin des Ronces, 
Théo for the leisurely walks along the Erdre, and Xavier for the board
game nights. 
% 
Meven, my thesis sibling, also played a huge part in making these three years
unforgettable --- thank you for being my protest comrade, for being my
drinking comrade, and for everything you taught me!

But now, the time has come for the next generation of PhD students. I wish all the best
to Martin, my favorite Toudou voice actor, to Pierre (Giraud) and his 
ability to know where is the closest piano at all times, and to Enzo and Hamza from
the next office --- hopefully you will get more time once I am not there to 
procrastinate with you anymore.
% 
And even though you are not yet PhD students, my wishes also apply to Yann and Arthur!

I am also grateful to the teaching team of the computer architecture course at the 
IUT de Nantes, and to the students I had the opportunity to teach. 
The teaching was certainly reciprocated, and I learned a lot from them. 

Finally, I would like to thank Steve for welcoming me in CMU, and all of the
people I had the occasion to meet there. Especially Anders, whom I had the
pleasure to work with.

Outside of the world of academia, I am lucky to have found excellent friends in
François, Vincent and Arthur, through ups and downs. They were such a great 
escape from working on my stubborn proofs!
% 
And during the monotonous times of lockdown and remote working, I could find a lot 
of fun and relief in a small, tight-knit online community --- I am of course thinking of the 
Mauve crew, les grands philosophes du quotidien.
I would also like to mention my ephemeral friends from Momiji, who showed 
me that even in the most unexpected place, I will always find kind and 
open-hearted people.

Last but not least, I want to thank Virginie Cornet-Berry and 
François Vellutini, two wonderful teachers whose passion for mathematics and love 
for transmission have accompanied me throughout my studies, as well as my family.
Thank you mom, thank you dad, thank you sis. Love you all.